---
title: 'My Teaching Script: intro to ggplot'
author: "Camila Vargas Poulsen"
date: '2022-07-12'
output: html_document
---

## Introduction
Hi everyone, for the last part of this session we are going to do a brief introduction to Data Visualization.


Data visualization is an important part of analysis, it as the way to see your data. 
Plotting your data is a good way to explore data, and the various relationships between variables. 

If you think about it, sometimes is much easier to learn from something that we can see rather than read.

Similarly how, sketching out the design for a house, communicates much more clearly than trying to describe it with words. The same thing is often true for data 

“Sketching” out our data by visualizing it is more impactful than simply describing the trends we find.

For today's lesson we are going to visualize data using `ggplot2` which is part of the tidyverse, this bundle of R packages that we have been using throughout the day where all packages share an underlying structure.

## Lesson's Goal
The goal of this lesson is 
- to introduce the basics of plotting data using the `ggplot2` package. 
- I will focus on the main components you need to create a plot using ggplot2. 
- We will barely scratch the surface of ggplot2's functionality.  
- But in the lesson's material you will find links to additional resources.

## Credits
This lesson is based on [OHI Data Science Trining](http://ohi-science.org/data-science-training/ggplot2.html), Chapter 5 Visualizing: `ggplot`


Let's get started...

## Plotting with `ggplot`
As I mentioned before the `ggplot2` package is the packages in the tidyverse that is responsible for visualization.


## Load tidyverse: `tidyverse`
At this point you should have the `tidyverse` installed in your machine
Remember that the `tidyverse` is a collection of R packages. So when we load the tidyverse to our R working space we are loading several packages. Within them `ggplot2`.

We use the `library()` function to load tidyverse.


```{r, messages = FALSE}
library(tidyverse) ## do this every time you restart R and need it
options(scipen=999)
```

You can see that all the packages form the `tidyverse` were load to your R work space (SHOW CONSOLE)

_If you get an error saying something like the package was not found, that meand you have to install the package._

## Load data

We will load the dataset we will be using for this tutorial into our work space directly from GitHub.com (I always find this GitHub functionality so cool!). You can read a url directly into your R environment with `read_csv` function.

_The data we will be using today is..._
We are going to use a National Park visitation dataset (from the National Park Service at https://irma.nps.gov/Stats/SSRSReports). Specifically, we are going to use data from California.

Read in the data using `read_csv`

I'll paste on the chat the url for you to read the data.

```{r load_data}
#National Parks in California
ca <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/ca.csv") 

```


## Explore the data
We are going to start by quickly exploring our dataset, to get familiarizes with it. Just be reading the data and creating an object in our environment we can see the we have 7 varibles (7 columnes) and 789 observations (rows).

Useful functions to explore your data:
```{r}
colnames(ca) ## prints all the column names
str(ca)
head(ca) ## first six rows
```


This dataframe is already in a *long* format where all rows are an observation and all columns are variables. Among the variables in `ca` are:

1. `region`, US region where park is located.

2. `visitors`, the annual visitation for each `year`


## Plotting with `ggplot2`

ggplot likes data in the ‘long’ format: i.e., a column for every dimension, and a row for every observation. Well structured data will save you lots of time when making figures with ggplot.

ggplot graphics are built step by step by adding new elements. Adding layers in this fashion allows for extensive flexibility and customization of plots.


One way of understanding ggplot2 is thinking about a figure in layers. This idea may be familiar to you if you have used image editing programs like Photoshop, Illustrator, or Inkscape. 
_<--! Remove this comment-->_
_Incorporate the layer concept when coding_

To build a ggplot we will always need:

- The `ggplot()` function,
- Indicate which data you want to plot using the `data` argument,
- define a coordinate system aesthetics (this means what to plot in the x axis and what to plot in the y axis) within the `aes()` function
- provide a geom (how do we want ggplot2 to plot our data). Graphical representation of the data in the plot (points,lines, bars)


**Aesthetic mappings describe how variables in the data are mapped to visual properties (aesthetics) of geoms. Aesthetic mappings can be set in ggplot() and in individual layers.**


```{r, eval=FALSE, purl=FALSE}
ggplot(data = ca,
       aes(x = year,
           y = visitors))+
  geom_point()
```


Notes:
- Anything you put in the `ggplot()` function can be seen by any geom layers
  that you add (i.e., these are universal plot settings). This includes the x and
  y axis you set up in `aes()`.
- You can also specify aesthetics for a given geom independently of the
  aesthetics defined globally in the `ggplot()` function.
- The `+` sign used to add layers must be placed at the end of each line containing
a layer. If, instead, the `+` sign is added in the line before the other layer,
**`ggplot2`** will not add the new layer and will return an error message.


This isn't necessarily a useful way to look at the data. We can distinguish each park by added the `color` argument to the `aes`:

```{r color-by-species, purl=FALSE}
ggplot(data = ca) +
    geom_point(aes(x = year, 
                   y = visitors, 
                   color = park_name))
```

## Customizing plots

Take a look at the [**`ggplot2`** cheat sheet](https://www.rstudio.com/wp-content/uploads/2016/11/ggplot2-cheatsheet-2.1.pdf), and think of ways you could improve the plot.

Now, let's capitalize the x and y axis labels and add a main title to the figure. I also like to remove that standard gray background using a different `theme`. Many themes come built into the `ggplot2` package. My preference is `theme_bw()` but once you start typing `theme_` a list of options will pop up. The last thing I'm going to do is remove the legend title.

```{r customize}
ggplot(data = ca) +
    geom_point(aes(x = year, 
                   y = visitors, 
                   color = park_name)) +
    labs(x = "Year",
       y = "Visitation",
       title = "California National Park Visitation") +
    theme_bw() +
    theme(legend.title=element_blank())
```


## Faceting

ggplot has a special technique called *faceting* that allows the user to split one plot into multiple plots based on data in the dataset. We will use it to make a plot of park visitation by state:

```{r first-facet, purl=FALSE}
ggplot(data = ca) +
    geom_point(aes(x = year, 
                   y = visitors,
                   color = park_name)) +
    facet_wrap(~park_name)
```



## Additional Material
There are many resources that will help you learn more. Here are some of the ones I use:   

- The official ggplot2 [cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf) is amazing!
- Preparing this teaching demonstration I came across this book
- _ADD A THIRD RESOURCE_




